---
title: "Lab 3"
author: "Team 28: Назар Пасічник, Анастасія Яблуновська, Марина Огінська"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

# Lab 4: Hypothesis testing

## Work Breakdown Structure

-   : Implementation of Problem 1 & 2
-   Яблуновська Анастасія: Implementation of Problem 3
-   : Implementation of Problem 4

# Problem 1

------------------------------------------------------------------------

# Problem 2

------------------------------------------------------------------------

# Problem 3

```{r}
id_number <- 28
n <- id_number

# Function to calculate a_k
a_k <- function(k, n) {
  frac_part <- (k * log(k^2 + n * pi)) %% 1
  return(frac_part)
}

a_data <- sapply(1:150, a_k, n = n)

x_vals <- qnorm(a_data[1:100])
y_vals <- qnorm(a_data[101:150])

```

## (a) KS Test for Normality

We test whether the sample $$
{ x_k }_{k=1}^{100}
$$ comes from a normal distribution with estimated parameters.
Hypotheses $$
H_0 : x_k \sim N(\hat{\mu}, \hat{\sigma}^2)
$$ $$
H_1 : x_k \not\sim N(\hat{\mu}, \hat{\sigma}^2)
$$ KS Statistic $$
D = \sup_x \left| F_n(x) - F(x) \right|
$$

```{r}
mu_hat <- mean(x_vals)
sigma_hat <- sd(x_vals)

ks_test_normal <- ks.test(x_vals, "pnorm",
                          mean = mu_hat, sd = sigma_hat)
ks_test_normal

if (ks_test_normal$p.value > 0.05) {
  cat("p-value =", ks_test_normal$p.value, "> 0.05. Fail to reject H0. Data appears normal.\n")
} else {
  cat("p-value =", ks_test_normal$p.value, "< 0.05. Reject H0. Data not normal.\n")
}

```


## (b) KS Test for Exponentiality of Absolute Values

We consider the transformed data: $$
{ |x_k| }_{k=1}^{100}
$$ We test whether they follow: $$
\text{Exp}(1)
$$ Hypotheses $$
H_0 : |x_k| \sim \text{Exp}(1)
$$ $$
H_1 : |x_k| \not\sim \text{Exp}(1)
$$

```{r}
abs_x <- abs(x_vals)

ks_test_exp <- ks.test(abs_x, "pexp", rate = 1)
ks_test_exp
if (ks_test_exp$p.value > 0.05) {
  cat("p-value =", ks_test_exp$p.value, "> 0.05. Fail to reject H0. |X| ~ Exp(1).\n")
} else {
  cat("p-value =", ks_test_exp$p.value, "< 0.05. Reject H0. |X| not Exp(1).\n")
}
```


## (c) Two-Sample KS Test for Equality of Distributions

We compare the samples $$
{ x_k }{k=1}^{100}
\qquad\text{and}\qquad
{ y\ell }_{\ell=1}^{50}
$$ Hypotheses $$
H_0 : F_X(x) = F_Y(x)\ \text{for all } x
$$ $$
H_1 : F_X(x) \neq F_Y(x)\ \text{for some } x
$$ KS Statistic $$
D = \sup_x \left| F_{X,n}(x) - F_{Y,m}(x) \right|
$$

```{r}
ks_test_two <- ks.test(x_vals, y_vals)
ks_test_two
if (ks_test_two$p.value > 0.05) {
  cat("p-value =", ks_test_two$p.value, "> 0.05. Fail to reject H0. Same distribution.\n")
} else {
  cat("p-value =", ks_test_two$p.value, "< 0.05. Reject H0. Different distributions.\n")
}
```

## Main Idea of KS Test

The Kolmogorov-Smirnov test compares the empirical cumulative
distribution function (ECDF) of sample data with a theoretical CDF. It
measures the maximum vertical distance D between them. If D exceeds a
critical value (or p-value \< α), we reject the hypothesis that data
follows the theoretical distribution.

------------------------------------------------------------------------

# Problem 4

------------------------------------------------------------------------

# Conclusions

## Problem 3

In this problem, we applied the Kolmogorov–Smirnov test to three
different hypotheses regarding the distributions of the generated
samples. First, we tested whether $$ {x_k}{k=1}^{100} $$ follows a
normal distribution with parameters estimated from the sample. Second,
we checked whether the transformed data $$ {|x_k|}{k=1}^{100} $$ is
exponentially distributed with rate $$ \lambda = 1. $$ Finally, we
compared the two samples
$$ {x_k}{k=1}^{100} \quad \text{and} \quad {y\ell}_{\ell=1}^{50} $$ to
determine whether they originate from the same distribution. Across all
three tests, the conclusions were based on comparing the p-value with
the significance level $$ \alpha = 0.05. $$ If the p-value exceeded this
threshold, we did not reject the null hypothesis; otherwise, we rejected
it. The KS test therefore allowed us to evaluate normality,
exponentiality, and the equivalence of distributions without assuming
any specific parametric form beyond the tested models.
