---
title: "Lab 4: Hypothesis Testing"
author: "Team 28: Назар Пасічник, Анастасія Яблуновська, Марина Огінська"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

# Lab 4: Hypothesis Testing

## Work Breakdown Structure

-   Назар Пасічник: Implementation of Problem 1 & 2
-   Марина Огінська: Implementation of Problem 3
-   : Implementation of Problem 4

# Problem 1

------------------------------------------------------------------------

**Task:** Test $H_0: \mu_1 = \mu_2$ vs. $H_1: \mu_1 \neq \mu_2$ with $\sigma_1^2 = \sigma_2^2 = 1$.

## Data Generation

First, we generate the data according to the lab instructions. The team ID n = 28 (sum of last digits).

```{r problem1_data}
# Team ID
n <- 28

# Function to calculate a_k
calc_a <- function(k, n) {
  value <- k * log(k^2 * n + pi)
  return(value - floor(value))  # fractional part
}

# Generate a_k for k = 1 to 100
a_data_x <- sapply(1:100, function(k) calc_a(k, n))

# Generate a_k for k = 101 to 150 (for Y)
a_data_y <- sapply(101:150, function(k) calc_a(k, n))

# Apply inverse CDF of standard normal to get X and Y
X <- qnorm(a_data_x)
Y <- qnorm(a_data_y)

# Sample sizes
n1 <- length(X)
n2 <- length(Y)

cat("Sample X: n1 =", n1, "\n")
cat("Sample Y: n2 =", n2, "\n")
```

## Test Selection and Assumptions

**Test Used:** Two-sample z-test for comparing means with known variances

**Why this test:**
- We are comparing means of two independent samples
- Both populations are normally distributed (by construction using qnorm)
- Variances are known and equal: $\sigma_1^2 = \sigma_2^2 = 1$ 
- When variances are known, we use the z-test; when variances are unknown, we would use the t-test
- This is a two-sided test since $H_1: \mu_1 \neq \mu_2$

**Test statistic:** $$Z = \frac{\bar{X} - \bar{Y}}{\sqrt{\frac{\sigma_1^2}{n_1} + \frac{\sigma_2^2}{n_2}}} = \frac{\bar{X} - \bar{Y}}{\sqrt{\frac{1}{100} + \frac{1}{50}}} \sim N(0,1) \text{ under } H_0$$

## Test Statistics and Calculations

```{r problem1_stats}
# Calculate sample statistics
mean_X <- mean(X)
mean_Y <- mean(Y)
var_X <- var(X)
var_Y <- var(Y)

cat("Sample statistics:\n")
cat("Mean of X:", round(mean_X, 4), "\n")
cat("Mean of Y:", round(mean_Y, 4), "\n")
cat("Variance of X:", round(var_X, 4), "\n")
cat("Variance of Y:", round(var_Y, 4), "\n")
cat("Difference in means:", round(mean_X - mean_Y, 4), "\n")

# Since variances are known to be 1, we use z-test formula
# Test statistic: Z = (X_bar - Y_bar) / sqrt(sigma^2/n1 + sigma^2/n2)
sigma_sq <- 1  # known variance
test_stat <- (mean_X - mean_Y) / sqrt(sigma_sq/n1 + sigma_sq/n2)

cat("\nTest statistic Z =", round(test_stat, 4), "\n")
```

## Rejection Region

For a two-sided test at significance level $\alpha = 0.05$:

**Rejection region:** $|Z| > z_{\alpha/2} = z_{0.025}$

```{r problem1_rejection}
alpha <- 0.05
z_critical <- qnorm(1 - alpha/2)

cat("Critical value: z_0.025 =", round(z_critical, 4), "\n")
cat("Rejection region: |Z| >", round(z_critical, 4), "\n")
cat("Observed |Z| =", round(abs(test_stat), 4), "\n")
```

## P-value Calculation

```{r problem1_pvalue}
# P-value for two-sided test
p_value <- 2 * (1 - pnorm(abs(test_stat)))

cat("P-value:", format(p_value, scientific = TRUE, digits = 4), "\n")
```

## Hypothesis Test Decision

```{r problem1_decision}
cat("Decision at alpha = 0.05:\n")
if (abs(test_stat) > z_critical) {
  cat("REJECT H0: |Z| =", round(abs(test_stat), 4), ">", round(z_critical, 4), "\n")
  cat("Conclusion: There is significant evidence that μ1 ≠ μ2\n")
} else {
  cat("FAIL TO REJECT H0: |Z| =", round(abs(test_stat), 4), "<=", round(z_critical, 4), "\n")
  cat("Conclusion: There is insufficient evidence to conclude that μ1 ≠ μ2\n")
}

cat("\nP-value interpretation:\n")
if (p_value < alpha) {
  cat("P-value =", format(p_value, scientific = TRUE, digits = 4), "< 0.05\n")
  cat("We reject H0 because the p-value is less than the significance level.\n")
  cat("This means the probability of observing such extreme data under H0 is very small.\n")
} else {
  cat("P-value =", format(p_value, scientific = TRUE, digits = 4), ">= 0.05\n")
  cat("We fail to reject H0 because the p-value exceeds the significance level.\n")
  cat("The data is consistent with H0.\n")
}
```

## Verification with R's Built-in Test

```{r problem1_verify}
# Using t.test with var.equal = TRUE (since we know variances are equal)
t_test_result <- t.test(X, Y, var.equal = TRUE, alternative = "two.sided")
print(t_test_result)
```

# Problem 2

------------------------------------------------------------------------

**Task:** Test $H_0: \sigma_1^2 = \sigma_2^2$ vs. $H_1: \sigma_1^2 > \sigma_2^2$ with $\mu_1$ and $\mu_2$ unknown.

## Test Selection and Assumptions

**Test Used:** F-test for equality of variances (one-sided)

**Why this test:**
- We are comparing variances of two independent samples
- Both populations are normally distributed (required assumption for F-test)
- The means are unknown but this doesn't affect the F-test
- This is a one-sided test since $H_1: \sigma_1^2 > \sigma_2^2$

**Theory:** Under $H_0$, the test statistic $F = S_1^2 / S_2^2$ follows an F-distribution with $(n_1-1, n_2-1)$ degrees of freedom, where $S_1^2$ and $S_2^2$ are sample variances.

## Data Preparation

We use the same data X and Y generated in Problem 1.

```{r problem2_data}
cat("Using data from Problem 1:\n")
cat("Sample X: n1 =", n1, "\n")
cat("Sample Y: n2 =", n2, "\n")
```

## Test Statistics and Calculations

```{r problem2_stats}
# Calculate sample variances
S1_sq <- var(X)
S2_sq <- var(Y)

cat("Sample variances:\n")
cat("S1^2 (variance of X):", round(S1_sq, 4), "\n")
cat("S2^2 (variance of Y):", round(S2_sq, 4), "\n")

# F-statistic
F_stat <- S1_sq / S2_sq

cat("\nF-statistic: F = S1^2 / S2^2 =", round(F_stat, 4), "\n")

# Degrees of freedom
df1 <- n1 - 1
df2 <- n2 - 1

cat("Degrees of freedom: df1 =", df1, ", df2 =", df2, "\n")
```

## Rejection Region

For a one-sided (right-tailed) test at significance level $\alpha = 0.05$:

**Rejection region:** $F > F_{\alpha}(df_1, df_2) = F_{0.05}(99, 49)$

We reject $H_0$ if the observed F-statistic exceeds the critical value from the F-distribution.

```{r problem2_rejection}
alpha <- 0.05
F_critical <- qf(1 - alpha, df1, df2)

cat("Critical value: F_0.05(", df1, ",", df2, ") =", round(F_critical, 4), "\n")
cat("Rejection region: F >", round(F_critical, 4), "\n")
cat("Observed F =", round(F_stat, 4), "\n")
```

## P-value Calculation

```{r problem2_pvalue}
# P-value for one-sided (right-tailed) test
p_value_f <- 1 - pf(F_stat, df1, df2)

cat("P-value:", format(p_value_f, scientific = FALSE, digits = 4), "\n")
```

## Hypothesis Test Decision

```{r problem2_decision}
cat("Decision at alpha = 0.05:\n")
if (F_stat > F_critical) {
  cat("REJECT H0: F =", round(F_stat, 4), ">", round(F_critical, 4), "\n")
  cat("Conclusion: There is significant evidence that σ1^2 > σ2^2\n")
  cat("The variance of population X is significantly greater than the variance of population Y.\n")
} else {
  cat("FAIL TO REJECT H0: F =", round(F_stat, 4), "<=", round(F_critical, 4), "\n")
  cat("Conclusion: There is insufficient evidence to conclude that σ1^2 > σ2^2\n")
  cat("The data does not provide strong evidence that variance of X exceeds variance of Y.\n")
}

cat("\nP-value interpretation:\n")
if (p_value_f < alpha) {
  cat("P-value =", format(p_value_f, scientific = FALSE, digits = 4), "< 0.05\n")
  cat("We reject H0 because the p-value is less than the significance level.\n")
  cat("This indicates strong evidence against the null hypothesis of equal variances.\n")
} else {
  cat("P-value =", format(p_value_f, scientific = FALSE, digits = 4), ">= 0.05\n")
  cat("We fail to reject H0 because the p-value exceeds the significance level.\n")
  cat("The observed difference in variances could reasonably occur by chance under H0.\n")
}
```

## Verification with R's Built-in Test

```{r problem2_verify}
# Using var.test for F-test
# alternative = "greater" tests if variance of X > variance of Y
var_test_result <- var.test(X, Y, alternative = "greater")
print(var_test_result)

cat("\nNote: R's var.test confirms our manual calculations.\n")
```

# Problem 3

------------------------------------------------------------------------

# Problem 4

------------------------------------------------------------------------

# Conclusions

## Problem 1: Testing Equality of Means

In Problem 1, we tested whether two independent samples X and Y have equal population means, assuming known and equal variances ($\sigma_1^2 = \sigma_2^2 = 1$). Using a two-sample z-test (or equivalently, a pooled t-test), we:

1. **Generated data** according to the specified transformation using team ID n = 28
2. **Calculated test statistic** Z based on the difference in sample means
3. **Determined rejection region** for a two-sided test at α = 0.05: |Z| > 1.96
4. **Computed p-value** to assess the strength of evidence against H₀

**Result:** Based on the test statistic and p-value, we made a decision to either reject or fail to reject H₀. The conclusion depends on whether the observed difference in sample means is statistically significant. If we rejected H₀, it indicates that the two populations have different means with 95% confidence. If we failed to reject H₀, the data does not provide sufficient evidence to conclude the means differ.

**Key insight:** Even though the theoretical variances are equal to 1, the sample variances may differ slightly due to random sampling. The test accounts for this variability through the standard error of the difference.

## Problem 2: Testing Equality of Variances

In Problem 2, we tested whether the variance of population X exceeds the variance of population Y using an F-test. The means were unknown, but this is accounted for in the F-test methodology. We:

1. **Applied F-test** as the appropriate method for comparing variances of normally distributed populations
2. **Calculated F-statistic** as the ratio S₁²/S₂² with degrees of freedom (99, 49)
3. **Determined rejection region** for a one-sided test at α = 0.05: F > F₀.₀₅(99, 49)
4. **Computed p-value** for the right-tailed test

**Result:** The decision to reject or fail to reject H₀ depends on whether the observed F-statistic exceeds the critical value. If we rejected H₀, this provides evidence that the first population has genuinely greater variability than the second. If we failed to reject H₀, the observed difference in sample variances could reasonably be attributed to random sampling variation.

**Key insight:** The F-test is sensitive to the normality assumption. Since our data was generated using the inverse normal CDF (qnorm), the normality assumption is satisfied by construction, making the F-test appropriate and reliable.

**Overall observation:** Both tests demonstrate the fundamental process of hypothesis testing: formulating hypotheses, selecting an appropriate test statistic, determining the rejection region, calculating the p-value, and making a decision based on the significance level. The p-value provides additional information about the strength of evidence against H₀, allowing for more nuanced interpretation than a simple reject/fail-to-reject decision.
